Model to study co-evolution of trade an culture

### 2015-03-31 article.tex
minor modifications
correct compile mistake
corrections from me on state of the art and model desctiption

\todo{and/or is ugly, look for another way}
This section will provide a brief review of the previous approaches to model trade and/or culture. While non-exhaustive, this review presents the mains trends of the previous work done.

@@

The first aspect of our model is focused on the dynamics of cultural change. They comprise the collection of processes that promote or inhibit the spread of information by social interaction within a population \cite[3]{boyd_origin_2005}. This evolutionary framework is being used by an increasing number of social scientists, fostering the development of transdisciplinary efforts designed to understand cultural evolution \cite{henrich_evolution_2003}. Archaeology has a rich tradition of evolutionary studies \cite{lycett_cultural_2015}, exploring concepts such as gene-culture coevolution \cite{burger_absence_2007}, diffusion processes \cite{fort_synthesis_2012}, phylogenetic analysis of material culture \cite{obrien_cladistics_2001} or detection of rates of change in the archaeological record \cite{premo_cultural_2014}.

A large percentage of these studies focus on the biases that affect the transmission of cultural traits, including the relevance of intrinsic traits and frequency dependance (e.g. conformism and anti-conformism). These analysis are often based in the genetic theory of neutral evolution \cite{neiman_stylistic_1995} and they take neutrality as a null-hypotheses (i.e. variation of traits is strictly explained by its frequency distribution in the population). There exist a large number of works using this model as the starting point to analyse patterns of cultural change in past societies \cite{lipo_neutralitystyle_2001,shennan_ceramic_2001,steele_ceramic_2010,kandler_nonequilibrium_2013,porcic_exploring_2014,crema_approximate_2014}.

@@

The second aspect of our model is focused on the dynamics produced by trade in ancient Mediterranean societes. Within the archaeological field a large number of works have been proposed to quantify the properties of the economical interactions in the ancient Mediterranean societies~\cite{hopkins_taxes_1980,temin_market_2001,terpstra_trade_2011,minc_monitoring_2006,temin_economy_2006,wilson_approaches_2009,scheidel_model_2007,kessler_organization_2007}. In the works of~\cite{temin_market_2001,temin_economy_2006,wilson_approaches_2009}, based on the analysis of archaeological evidence, the Roman economy has been classified as a type of market economy where a variety of small tightly connected markets had loose connections between them. This view of a market system in ancient societies has been further expended in~\cite{minc_monitoring_2006} by a model applicable to various cases and studied in a test case of the aztec economical structure. The reaming works attempted at defining more exactly which type market economy it was, i.e. was it close to ours or remote ?

The work of~\cite{scheidel_model_2007} proposes to analyse the factors affecting the growth of income and population in the Roman Italy. The analysis of historical evidence and historical documents lead the author to propose a model highlighting the key factors in the economical growth. More in link with the work conducted in this article~\cite{terpstra_trade_2011,kessler_organization_2007} have analysed the interactions between economy and culture transmission in roman empire. An analysis based on historical documents in~\cite{kessler_organization_2007} highlights the importance of economical and social institutions in order to perform an efficient trade of trains in the early Roman Empire. More precisely, the work conducted in~\cite{terpstra_trade_2011} proposes a micro-economic model analysing the possibility of trade under uncertainty in the roman empire. This work highlight the importance of reputation and the threat of expulsion from a trade network as mechanisms the overall efficiency of the trade network.

A limitation of the works cited above is the impossibility to test the conclusions reached by the authors. The models proposed reflects only the view of the authors on the data they have analyse. A new type of model would be necessary to go beyond the interpretation, and allow the verification of the conclusions reached. These models should be able to generate data which is then qualitatively compared to the real historical data. By this mean, if a factor is really important in the economical network, its removal from the model will lead to data qualitatively different from the historical data. The Multi-Agent Simulations (MAS) offer a way to produce this kind of model~\cite{lake_trends_2014}.

@@

Ultimately we aim to study the relation between cultural change and trade networks within the Roman Empire. Examples of questions that should be answered by our model are: What is the relation between distances in the trade network and cultural changes ? Did anti-conformism promote the adaptation of new products coming from remote provinces ? In order to answer these questions the model used has to take into account both cultural effects and trade influence. More precisely, the variations in these two aspects should be modelled and should influence each other.  Therefore, in order to answer the questions targeted we need a co-evolutionary model able to take into account both trade and cultural changes.

To the best of our knowledge two works have previously proposed frameworks that could be used to model the co-evolution of trade and culture~\cite{bentley_specialisation_2005,macmillan_agent-based_2008}. The first proposes a relatively simple model which is limited by 2 products, a fixed behaviour of the agents and a limited variation of cultural mechanisms. The second proposes a more complete model including a large number of goods (certain being essential and the others non-essential), but suffer from a greater complexity and the absence of the notion of trade networks. Moreover, both models propose new dynamics but these dynamics are not compared to the state of the art on the models for cultural change and trade.

The model presented in this work aims to offer a framework for the co-evolution of trades and culture which relies on a low number of parameters. In order to assess its validity we will reproduce results obtained in the literature and verify that the results obtained are conform to what is expected.

@@

In the model presented a set of agents are each in possession of a fixed number of goods and associate a value to each of these goods. Depending on the question studied, the value can reflect the interest of the agent for the good or it can be the price at which the agent evaluate this good. The agents are also endowed with limited storage abilities for each good that can be produced.

More formally the model is composed as follow. $r$ is the number of resources present in the environment. The whole population is defined by a vector of $m$ agents: 
$$ population = (a_i, \cdots ,a_m) $$

Each agent $i$ is defined by 2 vectors each of size $r$. The first correspond to the quantity of each good that the agent possesses: $$ quantity = (q^i_1,\cdots,q^i_r) $$

Each agent makes its own estimate on the price to attribute to a certain good. This price estimate is used for both selling and buying goods :
$$ values = (v^i_1,\cdots,p^i_r) $$

On top of these elements three processes are included : production, imitation and innovation. The \textit{production} process describe the creation of goods by the agent. In this article each agent is producing only one type of good. The type good produced by an agent is assigned to it at the beginning of the simulation and does not change. Once a good is produced it is added to the quantity vector of the agent. Within the \textit{imitation} process an agent can copy the value vector of another agent. The \textit{innovation} process reflects the possibility for an agent to associate new values for the goods. For this, an agent can select new elements for its value vector independently of the values of other agents.

This scheme allow us to have a model achieving the two main goals listed in section~\ref{sec:intro}. First, this model allows us to test various cultural transmission mechanisms as well as trade mechanisms. The experiments presented in this article show examples of implementations for both categories. Moreover, the fact that only three processes are modified allow comparisons between simulations with the same units. In this article the \textit{production} process will remain the same and only the \textit{imitation} and \textit{innovation} processes are changed. Second, few parameters are present in the default model which makes it easier to analyse and reuse by other researchers.

@@

\begin{table}
	\centering
	\begin{tabular}{l|cc}
N=250\\\hline
$\mu$&$\alpha$&(SD)\\\hline
0.004&1.50&(0.02) \\
0.008&1.52&(0.02)\\
0.016&1.55&(0.03)\\
0.032&1.59&(0.06)\\
0.064&1.78&(0.08)\\
0.128&1.99&(0.04)\\
	\end{tabular}
	\begin{tabular}{l|cc}
N=500\\\hline
$\mu$&$\alpha$&(SD)\\\hline
0.004&1.53&(0.03)
0.008&1.57&(0.01)
0.016&1.61&(0.04)
0.032&1.70&(0.05)
0.064&1.81&(0.10)
0.128&1.98&(0.02)
	\end{tabular}
	\caption{Mean \& SD from \cite{bentley_random_2004} }
	\label{tab:Bmualpha}
\end{table}

>>>>

This section will provide a brief review of the previous approaches to model trade and culture. While non-exhaustive, this review presents the mains trends of the previous work done.

@@

The first aspect of our model is focused on the dynamics of cultural change. They comprise the collection of processes that promotes or inhibits the spread of information by social interaction within a population \cite[3]{boyd_origin_2005}. An increasing number of social scientists are using an evolutionary framework to model these dynamics, thus fostering the development of transdisciplinary efforts designed to understand cultural evolution \cite{henrich_evolution_2003}. Archaeology has a rich tradition of evolutionary studies \cite{lycett_cultural_2015}, exploring concepts such as gene-culture coevolution \cite{burger_absence_2007}, diffusion processes \cite{fort_synthesis_2012}, phylogenetic analysis of material culture \cite{obrien_cladistics_2001} or detection of rates of change in the archaeological record \cite{premo_cultural_2014}.

A large percentage of these studies focus on the biases that affect the transmission of cultural traits, including the relevance of intrinsic traits and frequency dependence (e.g. conformism and anti-conformism). 
\jm{to keep it simple, can we say ``evolutionary theory" instead of ``genetic theory" ?}
These analysis are often based on the genetic theory of neutral evolution \cite{neiman_stylistic_1995} and they take neutrality as a null-hypotheses, i.e. variations of traits are strictly explained by their frequency distribution in the population. There exist a large number of works using this model as the starting point to analyse patterns of cultural change in past societies \cite{lipo_neutralitystyle_2001,shennan_ceramic_2001,steele_ceramic_2010,kandler_nonequilibrium_2013,porcic_exploring_2014,crema_approximate_2014}.

@@

Previous historical and archaeological studies to model trade in the ancient Mediterranean societies have focus on descriptive approaches ~\cite{hopkins_taxes_1980,temin_market_2001,terpstra_trade_2011,temin_economy_2006,wilson_approaches_2009,scheidel_model_2007,kessler_organization_2007}. 
These works have conclude that the Roman economy was a type of market economy where a variety of small tightly connected markets had loose connections between them~\cite{temin_market_2001,temin_economy_2006,wilson_approaches_2009}. 
Based on these conclusions, further works have attempt to define more exactly which type market economy was in use, i.e. was it close to our market economy or remote ?

Among the various works done with this objective, few are studying the interactions between economy and cultural change in Roman empire~\cite{terpstra_trade_2011,kessler_organization_2007,scheidel_model_2007}. 
An analysis based on historical documents has revealed the importance of economical and social institutions for the efficiency of trades of grains in the early Roman Empire~\cite{kessler_organization_2007}. The same type of analysis has also revealed the importance of a number of key factors () for the growth of income and population in the Roman Italy~\cite{scheidel_model_2007}. Finally, a micro-economic model has determined the importance of reputation and threat of expulsion from a trade network as mechanisms to maintain the overall efficiency of the trade network~\cite{terpstra_trade_2011}. These works therefore reveal the strong relation between trade and cultural change in the Roman empire.

However, a limitation of the previous works is the impossibility to test the conclusions reached by the authors. The models proposed reflects only the view of the authors on the data they have analyse. A new type of model would be necessary to go beyond the interpretation, and allow the verification of the conclusions reached. These models should be able to generate data which is then qualitatively compared to the real historical data. By this mean, if a factor is really important in the economical network, its removal from the model will lead to data qualitatively different from the historical data. The Multi-Agent Simulations (MAS) offer a way to produce this kind of model~\cite{lake_trends_2014}.

@@

Ultimately, we aim to create a model to study the relation between cultural change and trade networks within the Roman Empire. Examples of questions that could be asked to our model are: What is the relation between distances in the trade network and cultural changes ? Did anti-conformism promote the adaptation of new products coming from remote provinces ? In order to answer these questions the model used has to take into account both cultural effects and trade influence. More precisely, the variations in these two aspects should be modelled and should influence each other.  Therefore, in order to answer the questions targeted we need a co-evolutionary model able to take into account both trade and cultural changes.

To the best of our knowledge two works have previously proposed frameworks that could be used to model the co-evolution of trade and culture~\cite{bentley_specialisation_2005,macmillan_agent-based_2008}. The first proposes a relatively simple model which is limited by 2 products, a fixed behaviour of the agents and a limited variation of cultural mechanisms. The second proposes a more complete model including a large number of goods (certain being essential and the others non-essential), but suffer from a greater complexity and the absence of the notion of trade networks. Moreover, both models propose new dynamics but these dynamics are not compared to previous results obtained in the study of cultural evolution and the study of trading mechanisms.

The model presented in this work aims to offer a framework for the co-evolution of trades and culture which relies on a low number of parameters. It is not within the scope of this article to answer the questions asked at the beginning of this section as they require specific implementations of the model proposed. However, the article will focus on the presentation of the framework, its justification and more importantly its validation by reproducing previous results obtained on cultural change and trade mechanisms.

@@

The model proposed is a Multi-Agent Based model including agents and products. It revolves around the value that agents associate to products through cultural change mechanisms, trade mechanisms or both. Depending on the question studied, the value can reflect the interest of the agent for the good or it can be the price at which the agent evaluate this good. The agents are also endowed with limited storage abilities for each good that can be produced.

More formally the model is composed of a population $Pop$ of $m$ agents, each defined by 2 vectors of size $r$. The first corresponds to the quantity of each good that the agent $i$ possesses: $$ Q^i = (q^i_1,\cdots,q^i_r) $$

where $Q^i$ is the total list of possessions of agent $i$, and $q^i_j$ is the number of goods of type $j$ that agent $i$ posses.

The second vector reflects the estimation of the value of a product that an agent $i$ makes.
$$ V^i = (v^i_1,\cdots,v^i_r) $$
where $V^i$ is the total list of estimated values of agent $i$, and $v^i_j$ is the value that agent $i$ associates to the goods of type $j$.

On top of these elements three processes are included : \textit{production}, \textit{imitation} and \textit{innovation}. The \textit{production} process describe the creation of goods by the agent. Once a good is produced it is added to the quantity vector of the agent. Within the \textit{imitation} process an agent can copy the value vector of another agent. The \textit{innovation} process reflects the possibility for an agent to associate new values for the goods. For this, an agent can select new elements for its value vector independently of the values of other agents.

This scheme allow us to have a model achieving the two main goals listed in section~\ref{sec:intro}. The first goal is to be able to implement and compare various cultural transmission mechanisms as well as trade mechanisms within the same framework. We argue that this is possible in our framework by modifying only the three key processes i.e. production, imitation and innovation. Moreover, thanks to the use of a unique $V^i$ vector, it is possible to compare the outcomes of different combinations of production, imitation and innovation mechanisms. The second goal is to produce a model composed of a low number of parameter. Our model is composed of only three processes and two vectors which is significantly less complex than previous models.

\todo{check that it's integrated later: In this article each agent is producing only one type of good. The type good produced by an agent is assigned to it at the beginning of the simulation and does not change.}

\todo{check that it's integrated later: 
In this article the \textit{production} process will remain the same and only the \textit{imitation} and \textit{innovation} processes are changed.}

@@

\begin{table}
	\centering
	\begin{tabular}{l|cc}
N=250\\\hline
$\mu$&$\alpha$&(SD)\\\hline
0.004&1.50&(0.02) \\
0.008&1.52&(0.02)\\
0.016&1.55&(0.03)\\
0.032&1.59&(0.06)\\
0.064&1.78&(0.08)\\
0.128&1.99&(0.04)\\
	\end{tabular}
	\begin{tabular}{l|cc}
N=500\\\hline
$\mu$&$\alpha$&(SD)\\\hline
0.004&1.53&(0.03)\\
0.008&1.57&(0.01)\\
0.016&1.61&(0.04)\\
0.032&1.70&(0.05)\\
0.064&1.81&(0.10)\\
0.128&1.98&(0.02)\\
	\end{tabular}
	\caption{Mean \& SD of $\alpha$ parameter }
	\label{tab:Bmualpha}
\end{table}
